Measuring contrast processing in the visual system using SSVEP

### 2024-04-06 review.qmd
Implemented figure referencing

Contrast is the currency of the early visual system. Measuring the way that the computations underlying contrast processing depend on factors such as spatial and temporal frequency, age, eccentricity, chromaticity and the presence of other stimuli has been a focus of vision science for at least 100 years. One of the most productive experimental approaches in this field has been the use of the 'steady-state visually-evoked potential' (SSVEP): a technique where contrast modulating inputs are 'frequency tagged' (presented at well-defined frequencies and phases) and the electrical signals that they generate in the brain are analyzed in the temporal frequency domain. SSVEPs have several advantages over conventional measures of visually-evoked responses: they have relatively unambiguous ouput measures, a high SNR, and they allow us to analyze interactions between stimulus components using a convenient mathematical framework. Here we describe how SSVEPs have been used to study visual contrast over the past 70 years [@Dawson1954]. Because our thinking about SSVEPs is well-described by simple mathematical models, we embed code that illustrates key steps in the modelling and analysis. This document can therefore be used both as a review of the use of SSVEP in measuring human contrast processing, and as an interactive learning aid.

@@

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal, although careful analysis of individual VEPs reveals multiple generators throughout visual cortex [@Di-Russo2007] @Hillebrand. The flickering stimulus entrains neural responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker tends to produce more complex spectral harmonics than sine-wave flicker.

@@

These contrast definitions are appropriate both to photometric measures of stimulus contrast (for example, luminance; @Lennie1993) and also to definitions based on cone excitations [@Derrington1984, @MacLeod1979] which are more common in work on chromatic processing.

Although its definition as a stimulus is simple, the computations that underlie contrast processing in the brain have been the subject of intense research for many decades. The neural code for contrast, even in the earliest parts of visual cortex, is not simply a linear transform of the contrast at the retina - instead, contrast signals undergo a cascade of nonlinear processing stages that, broadly, attempt to normalise the output relative to the spatiotemporal environment. This normalization, achieved through a computation called 'contrast gain control' [@Foley1994; @Heeger1992; @Carandini2011] maximises the sensitivity of the visual system by making optimal use of neuronal bandwidth. As an example, a grating placed at the centre of a low-contrast background typically appears more intense than the same grating when superimposed on a high contrast background.

```{python}
#| label: Perceptual gain control
#| fig-cap: The perceived contrast of a stimulus depends on its context. A high contrast surround reduces the apparent contrast of the central 'probe' region"
#| fig-cap-location: bottom

@@

A significant part of research into contrast processing is concerned with how these normalization mechanisms depend on colour [@Chen2000], orientation [@Foley1994], eye of origin [@Legge1979; @Baker2007], spatial and temporal frequency [@Meese2009], location [@Petrov2005; @Tadin2003; @Polat1993], age [@Betts2005], and the presence of neurological disorders [@Porciatti2000; @Tsai2011]. The SSVEP has proven to be invaliable in this research because it provides an objective readout of contrast representation at different parts of the visual system and allows us to 'tag' the probe and background at separate frequencies.

Because it provides a direct read-out of neural population activity, the SSVEP signal can reveal key features of neural signal transduction. For example, by varying the peak stimulus contrast parametrically, a 'contrast vs response' function can be measured - where the 'response' is typically defined as the amplitude of the SSVEP frequency component at the stimulus frequency or a low multiple thereof. This corresponds closely to similar functions measured in single unit studies, or those measuring local field potentials invasively in the cortex [@Shapley1980]. However the SSVEP has the advantage that it is non-invasive, and so can be measured in awake, behaving human participants.

@@

Neurons have a limited dynamic range yet they can transmit information about visual stimuli that span many orders of magnitude. In the domain of contrast, to some extent this is accomplished at a population level - individual neurons typically implement a non-linear, sigmoidal contrast vs response (CRF) transducer [@Albrecht1982; @Tolhurst1981] and different neurons exhibit peak sensitivity (defined as the maximum slope of the function) at different contrast levels [@Carandini1994; @Carandini1998; @Busse2009]. A neuronal population will therefore span a sensitivity range greater than any individual member.

@@

Previous figures showed how a single-input SSVEP might generate a single set of peaks in the frequency domain response which, in turn, can be used as a read-out of neuronal activity.

Here, we show the effect of adding a second contrast component: in general, the two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms ---  is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan1988]

First, we implement a simple non-linear transducer with a single input to illustrate the typical signoidal shape of a contrast response function.
(Figure xx)

Next, we show that adding a second component which contributes to the gain pool of the first, results in a complex pattern of intermodulation terms.

Figures:

```{python}
#| label: plotsinewaves
#| fig-cap: "a) A typical SSVEP setup: a screen, a grating, a flicker rate: b) The responses from a single electrode (time domain) c) Frequency domain"
#| fig-cap-location: bottom

@@

If the two inputs are simply added together, the representation of the resulting signal in the Fourier domain is just the linear sum of the two independent signals. However, it is important to take physiology into account. If we ignore cells sensitive to contrast transients, contrast is represented as a combination of rectified signals from 'on' and 'off'-sensitive cells (broadly, cells that code positive or negative changes in spatiotemporal luminance). To model the population response of these cells to a contrast-reversing luminance sine wave, we therefore introduce an initial full-wave rectification stage that simulates the outputs of two, equal populations of on- and off-polarity detectors. Once this is done, we see that the Fourier spectrum of the resulting signal contains power only at the second harmonics and above. It is possible to generate first-hamonic responses in contrast-response SSVEPs but only when the inputs are modulated in an 'on/off' manner rather than contrast reversed.

```{python}
#| label: plotspectra
#| fig-cap: "Temporal frequency representations of combining two inputs at frequencies F1=5Hz and F2=7Hz combination. A pure sine-wave input would generate responses only at F1 and F2. However, the rectification that accompanies early visual processing converts this contrast reversal into a more complex frequency-domain representation with power predominantly at the second harmonics 2F1 and 2F2"
#| fig-cap-location: bottom

@@

One of the simplest nonlinearities is the response function that describes a cell's response to different levels of contrast. In Figure xxx this is modeled as described above by a hyperbolic ratio function resulting in a saturating non-linearity:

@@

where ${R_{\mathrm{max}}}$ describes the maximum response level, $C_{\mathrm{in}}$ is the input contrast, $C_{50}$ is the 'semi-saturation constant' (the point at which the reponse is at half-maximum) and *n* controls the steepness of the curve.

```{python}
#| label: ContrastResponseFunction
#| fig-cap: "Non-linear contrast response function resulting from the population average of two full-wave rectified signals passing through a non-linear transducer. Note the 'supersaturation' at high contrast levels."
#| fig-cap-location: bottom

@@

The hyperbolic ratio function is monotonic but the CRF resulting from measuring the amplitude of the 2F component contains a slight roll-off at high input contrasts. This results from the distortion of the input sine waves at high contrast due to a combination of the full-wave rectification and saturating non-linearity. Power at other harmonics increases and the total power generated by the input is monotonic. This roll-off is often seen in experimental data and has been referred to as 'supersaturation' [@Tyler1985; @Peirce2007].

To illustrate the effect of contrast gain control [@Heeger1992], we include a second component that contributes to the gain pool of the first. At a single pair of (matched) contrast levels, we see a complex pattern of intermodulation terms in the FT.

```{python}
#| label: Intermodulation
#| fig-cap: "Non-linear combination of two independent contrast modulations at F1 and F2 results in 'intermodulation' terms that appear at low-order sums and differences of those frequencies"
#| fig-cap-location: bottom

@@

The complexity of even a simple simulation of the frequency-domain signal arising from non-linear interactions is intriguing. Presumably, the signal measured from early visual cortex contains is the result of a cascade of nonlinear retinal and cortical operations up to that point and therefore contains a 'signature' of the computational nature, order and parameters of those operations - including the shape of the transducer functions and the computations involved in signal combination. In principle, that information could be recovered from the SSVEP signal.

@@

An early use of the SSVEP was to provide an objective estimate of spatial contrast sensitivity in infants, without requiring behavioural responses.

In well-motivated adults, psychophysical measurements of contrast sensitivity remain the gold standard. However, it is difficult and time consuming to obtain reliable psychophysical data from infants. In these cases, SSVEP measurements represent a fast and efficient method for measuring low-level visual responses [@Norcia1990; @Tyler1979; @Braddick1986] and the high SNR of SSVEP means that infants need only look at the screen for short periods of time.

Because SSVEP response at detection threshold are very small, estimating this threshold is achieved by measuring the contrast response function at relatively high levels, and extrapolating back along the function (either contrast vs response measured at a constant spatial frequency or spatial frequency vs response at a constant contrast level) to estimate its intercept with the x-axis. This contrast level was shown to correspond approximately with psychophysically measured detection thresholds [@Norcia1986].

Measuring the SSVEP at multiple levels was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, spatial frequency or some other property, throughout a trial [@Tyler1979]. To avoid hysteresis effects, the sweep is sometimes conducted both up and down in the same experiment [@Tyler1979, @Norcia1985a; @Norcia1990].

This general approach has revealed much about the development of visual abilities in infants [@Harris1976; @Braddick1986; @Atkinson1979]. In general, SSVEP measurements of infant vision have revealed that contrast sensitivity for both achromatic and chromatic contrast as well as stereoscopic depth perception develops earlier than had been supposed previously based on behavioural readouts [@Dobson1978; @Norcia1985b] with both chromatic and achromatic contrast detection reaching near-adult levels by around six months. Spatial acuity as measured by SSVEP reaches adult levels more slowly, but near-adult levels are recorded around 1 year [@Norcia1985b; @Hamer1989] compared to around 6-7 years with behavioural measures [@Ellemberg1999; @Atkinson1983]. At least some of this difference is likely due to the relative objectivity and high SNR of the SSVEP technique compared to other methods such as preferential looking, which require careful measurement of the infant's gaze direction. However it should be noted that other groups have reported electrophysiological correlates of visual acuity that more closely match the behavioural measures [@De-Vries-Khoe1982].

The SSVEP technique has also been used to study the development of the contrast gain control mechanisms described in the previous section.[@Candy2001; @Pei2017]. Although contrast gain control is measurable in infants as young as six weeks old [@Candy2001; @Morrone1986; @Skoczenski1998], its development appears to be slower, with adult levels being reached at approximately 11 years [@Pei2017].

@@

#| label: SweepVEP Deteection threshold estimate
#| fig-cap: "Sweep VEP simulation showing how a contrast detection threshold can be estimated from sweep VEP data measured at higher contrasts"
#| fig-cap-location: bottom

@@

The sweep VEP (really, a sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. In particular, because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004; @Hoffmann2017; @Bach2008; @Bach2020].

>>>>

Contrast is the currency of the early visual system. Measuring the way that the computations underlying contrast processing depend on factors such as spatial and temporal frequency, age, eccentricity, chromaticity and the presence of other stimuli has been a focus of vision science for at least 100 years. One of the most productive experimental approaches in this field has been the use of the 'steady-state visually-evoked potential' (SSVEP): a technique where contrast modulating inputs are 'frequency tagged' (presented at well-defined frequencies and phases) and the electrical signals that they generate in the brain are analyzed in the temporal frequency domain. SSVEPs have several advantages over conventional measures of visually-evoked responses: they have relatively unambiguous ouput measures, a high SNR, and they allow us to analyze interactions between stimulus components using a convenient mathematical framework. Here we describe how SSVEPs have been used to study visual contrast over the past 70 years [@Dawson1954]. Because our thinking about SSVEPs is well-described by simple mathematical models, we embed code that illustrates key steps in the modelling and analysis. This paper can therefore be used both as a review of the use of SSVEP in measuring human contrast processing, and as an interactive learning aid.

@@

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal, although careful analysis of individual VEPs reveals multiple generators throughout visual cortex [@Di-Russo2007] @Hillebrand. The flickering stimulus entrains neural population responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker tends to produce more complex spectral harmonics than sine-wave flicker.

@@

These contrast definitions are appropriate both to photometric measures of stimulus contrast (for example, luminance; @Lennie1993) and also to definitions based on cone excitations [@Derrington1984; @MacLeod1979] which are more common in work on chromatic processing.

Although its mathematical definition is straightforward, the computations that underlie contrast processing in the brain have been the subject of intense research for many decades. The neural code for contrast, even in the earliest parts of visual cortex, is not simply a linear transform of the contrast at the retina - instead, contrast signals undergo a cascade of nonlinear processing stages that, broadly, attempt to normalise the output relative to the spatiotemporal environment. This normalization, achieved through a computation called 'contrast gain control' [@Foley1994; @Heeger1992; @Carandini2011] maximises the sensitivity of the visual system by making optimal use of neuronal bandwidth. As an example, a grating placed at the centre of a low-contrast background typically appears more intense than the same grating when superimposed on a high contrast background (see @fig-centresurround).

```{python}
#| label: fig-centresurround
#| fig-cap: The perceived contrast of a stimulus depends on its context. A high contrast surround reduces the apparent contrast of the central 'probe' region"
#| fig-cap-location: bottom

@@

A significant body of research into contrast processing is concerned with how these normalization mechanisms depend on colour [@Chen2000], orientation [@Foley1994], eye of origin [@Legge1979; @Baker2007], spatial and temporal frequency [@Meese2009], location [@Petrov2005; @Tadin2003; @Polat1993], age [@Betts2005], and the presence of neurological disorders [@Porciatti2000; @Tsai2011]. The SSVEP has proven to be invaliable in this research because it provides an objective readout of contrast representation at different stages of the visual system, and allows us to 'tag' the probe and background at separate frequencies.

Because it provides a direct read-out of neural population activity, the SSVEP signal can reveal key features of neural signal transduction. For example, by varying the peak stimulus contrast parametrically, a 'contrast vs response' function can be measured - where the 'response' is typically defined as the amplitude of the SSVEP frequency component at the stimulus frequency or a low multiple thereof. This corresponds closely to similar functions reported by invasive studies measuring single unit activity or local field potentials in the cortex [@Shapley1980; @Morrone1982]. However the SSVEP has the advantage that it is non-invasive, and so can be measured in awake, behaving human participants.

@@

Neurons have a limited dynamic range, yet they can transmit information about visual stimuli that span many orders of magnitude. In the domain of contrast, to some extent this is accomplished at a population level - individual neurons typically implement a non-linear, sigmoidal contrast vs response (CRF) transducer [@Albrecht1982; @Tolhurst1981] and different neurons exhibit peak sensitivity (defined as the maximum slope of the function) at different contrast levels [@Carandini1994; @Carandini1998; @Busse2009]. A neuronal population will therefore span a sensitivity range greater than any individual member.

@@

Previous figures showed how a single-input SSVEP might generate a single set of peaks in the frequency domain response which, in turn, can be used as a read-out of neuronal activity. In @fig-plotspectra, we show the effect of adding a second contrast component: the two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms ---  is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan1988]

First, we implement a simple non-linear transducer with a single input to illustrate the typical sigmoidal shape of a contrast response function [@fig-ContrastResponseFunction].

Next, we show that adding a second component which contributes to the gain pool of the first, results in a complex pattern of intermodulation terms [@fig-plotspectra].

```{python}
#| label: fig-plotsinewaves
#| fig-cap: "a) A typical SSVEP setup: a screen, a grating, a flicker rate: b) The responses from a single electrode (time domain) c) Frequency domain"
#| fig-cap-location: bottom

@@

If the two inputs are simply added together, the representation of the resulting signal in the Fourier domain is just the linear sum of the two independent signals (first two panels of @fig-plotspectra). However, it is important to take physiology into account. If we ignore cells sensitive to contrast transients, contrast is represented as a combination of rectified signals from 'on' and 'off'-sensitive cells (broadly, cells that code positive or negative changes in spatiotemporal luminance). To model the population response of these cells to a contrast-reversing luminance sine wave, we therefore introduce an initial full-wave rectification stage that simulates the outputs of two, equal populations of on- and off-polarity detectors. Once this is done, we see that the Fourier spectrum of the resulting signal contains power only at the second harmonics and above (final two panels of @fig-plotspectra). It is possible to generate first-hamonic responses in contrast-response SSVEPs, but only when the inputs are modulated in an 'on/off' manner rather than contrast reversed.

```{python}
#| label: fig-plotspectra
#| fig-cap: "Temporal frequency representations of combining two inputs at frequencies F1=5Hz and F2=7Hz combination. A pure sine-wave input would generate responses only at F1 and F2. However, the rectification that accompanies early visual processing converts this contrast reversal into a more complex frequency-domain representation with power predominantly at the second harmonics 2F1 and 2F2"
#| fig-cap-location: bottom

@@

One of the simplest nonlinearities is the function that describes a cell's response to different levels of contrast. In @fig-ContrastResponseFunction this is modeled as described above by a hyperbolic ratio function resulting in a saturating non-linearity:

@@

where ${R_{\mathrm{max}}}$ describes the maximum response level, $C_{\mathrm{in}}$ is the input contrast, $C_{50}$ is the 'semi-saturation constant' (the point at which the response is at half-maximum) and *n* controls the steepness of the curve.

```{python}
#| label: fig-ContrastResponseFunction
#| fig-cap: "Non-linear contrast response function resulting from the population average of two full-wave rectified signals passing through a non-linear transducer. Note the 'supersaturation' at high contrast levels."
#| fig-cap-location: bottom

@@

The hyperbolic ratio function is monotonic but the CRF resulting from measuring the amplitude of the 2F component contains a slight roll-off at high input contrasts. This results from the distortion of the input sine waves at high contrast due to a combination of the full-wave rectification and saturating non-linearity. Power at other harmonics increases, and the total power generated by the input is monotonic. This roll-off is often seen in experimental data and has been referred to as 'supersaturation' [@Tyler1985; @Peirce2007].

To illustrate the effect of contrast gain control [@Heeger1992], we include a second component that contributes to the gain pool of the first. At a single pair of (matched) contrast levels, we see a complex pattern of intermodulation terms in the FT [@fig-Intermodulation].

```{python}
#| label: fig-Intermodulation
#| fig-cap: "Non-linear combination of two independent contrast modulations at F1 and F2 results in 'intermodulation' terms that appear at low-order sums and differences of those frequencies"
#| fig-cap-location: bottom

@@

The complexity of even a simple simulation of the frequency-domain signal arising from non-linear interactions is intriguing. Presumably, the signal measured from early visual cortex contains is the result of a cascade of nonlinear retinal and cortical operations up to that point. It therefore contains a 'signature' of the computational nature, order and parameters of those operations - including the shape of the transducer functions and the computations involved in signal combination. In principle, that information could be recovered from the SSVEP signal.

@@

An early use of the SSVEP was to provide an objective estimate of spatial contrast sensitivity in infants, without requiring behavioural responses. In well-motivated adults, psychophysical measurements of contrast sensitivity remain the gold standard. However, it is difficult and time consuming to obtain reliable psychophysical data from infants. In these cases, SSVEP measurements represent a fast and efficient method for measuring low-level visual responses [@Norcia1990; @Tyler1979; @Braddick1986] and the high SNR of SSVEP means that infants need only look at the screen for short periods of time.

Because SSVEP responses at detection threshold are very small, estimating threshold is achieved by measuring the contrast response function at relatively high levels, and extrapolating back along the function (either contrast vs response measured at a constant spatial frequency or spatial frequency vs response at a constant contrast level) to estimate its intercept with the x-axis (see @fig-SweepVEP). This contrast level was shown to correspond approximately with psychophysically measured detection thresholds [@Norcia1986].

@@

#| label: fig-SweepVEP
#| fig-cap: "Sweep VEP simulation showing how a contrast detection threshold can be estimated from sweep VEP data measured at higher contrasts"
#| fig-cap-location: bottom

@@

Measuring the SSVEP at multiple levels was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, spatial frequency or some other property, throughout a trial [@Tyler1979]. To avoid hysteresis effects, the sweep is sometimes conducted both up and down in the same experiment [@Tyler1979, @Norcia1985a; @Norcia1990]. The sweep VEP (really, a sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. In particular, because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004; @Hoffmann2017; @Bach2008; @Bach2020].

This approach has revealed much about the development of visual abilities in infants [@Harris1976; @Braddick1986; @Atkinson1979]. In general, SSVEP measurements of infant vision have revealed that contrast sensitivity for both achromatic and chromatic contrast as well as stereoscopic depth perception develops earlier than had been supposed previously based on behavioural readouts [@Dobson1978; @Norcia1985b] with both chromatic and achromatic contrast detection reaching near-adult levels by around six months. Spatial acuity as measured by SSVEP reaches adult levels more slowly, but near-adult levels are recorded around 1 year [@Norcia1985b; @Hamer1989] compared to around 6-7 years with behavioural measures [@Ellemberg1999; @Atkinson1983]. At least some of this difference is likely due to the relative objectivity and high SNR of the SSVEP technique compared to other methods such as preferential looking, which require careful measurement of the infant's gaze direction. However it should be noted that other groups have reported electrophysiological correlates of visual acuity that more closely match the behavioural measures [@De-Vries-Khoe1982].

The SSVEP technique has also been used to study the development of the contrast gain control mechanisms described in the previous section [@Candy2001; @Pei2017]. Although contrast gain control is measurable in infants as young as six weeks old [@Candy2001; @Morrone1986; @Skoczenski1998], its development appears to be slower, with adult levels being reached at approximately 11 years [@Pei2017].
